---
title: "M8L2 Homework Assignment"
author: "Joshua Conte"
date: "November 26, 2017"
output: 
  pdf_document:
    fig_caption: true
header-includes:
    - \usepackage{fancyhdr}
    - \pagestyle{fancy}
    - \fancyhead[CO,CE]{DA5030}
    - \fancyfoot[CO,CE]{Assignment M8L2 by Joshua Conte}
    - \fancyfoot[LE,RO]{\thepage}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
&nbsp;


\clearpage
\setcounter{secnumdepth}{5}
\setcounter{tocdepth}{5}
\pagebreak

# M8L2 Homework Assignment
R studio was configured with the following parameters before beginning the project:
```{r Configure RStudio, warning = FALSE, message = FALSE}
# clears the console in RStudio
cat("\014") 

# clears environment
rm(list = ls())  

# Load required packages
library(RCurl)
library(plyr)
library(forecast)
```

## Load the Data

I will be using an output file from my project HGMD_final_CHD2.csv. This data contains the publication year for mutations in the CHD2 gene. I am going to use this data to predict futur publications for mutations in this gene.

```{r}
dfHGMD_final <- read.csv2(
        "HGMD_final_CHD2.csv",
        sep = ",",
        stringsAsFactors = FALSE
      )
```

### Formating the data
Inorder to analyze the data I need numeric data. I will make new columns with Consequence, Variantclass, Reported.phenotype, Mutation.type, Overall.consequence, and year converted to numeric.

For Consequence the data will change to:
Create column and assign a Consequence with a num
Ser112Term = 1, Arg121Term = 2, Pro218Leu = 3, Arg466Term = 4, Leu823Pro = 5,
Asp856Gly = 6, Gln906Term = 7, Gly1174Asp = 8, Arg1313Gly = 9, Arg1345Gln = 10,
Ser1406Gly = 11, Trp1534Cys = 12, Arg1637Term = 13, Gln1641Term = 14, 
Trp1657Term = 15, Arg1679Term = 16, c.390C>T = 17, c.1502+1G>A = 18, 
c.1719+5G>A = 19, c.1810-2A>C = 20, c.1552delC = 21, c.1809+1delG = 22,
c.1880_1883delCTTT = 23, c.2895_2898delAGAA = 24, c.3734delA = 25, 
c.4233_4236delAGAA = 26, c.4256_4274del19 = 27, c.3787dupG = 28, c.4173dupA = 29,
c.4949dupG = 30, c.5094dupG = 31
```{r Consequence.num}
# Make Consequence numeric
dfHGMD_final$Consequence.num <- 1:nrow(dfHGMD_final) 
```

For Reported Phenotype the data will change to:
Autism = 1
Epilepsy = 2
Eyelid myoclonia = 3
Intellectual disability = 4
```{r Reported.phenotype.num}
# Make Reported.phenotype numeric
# Assign number to Reported.phenotype
# Create column
dfHGMD_final$Reported.phenotype.num <-
  dfHGMD_final$Reported.phenotype

# Use regex to make data numeric
# Autism = 1
dfHGMD_final$Reported.phenotype.num <-
  gsub("^Autism.*",
       dfHGMD_final$Reported.phenotype.num,
       replacement = "1")

# Epilepsy = 2
dfHGMD_final$Reported.phenotype.num <-
  gsub("^Epilep.*|^Dravet.*|^Lennox-Gastaut.*",
       dfHGMD_final$Reported.phenotype.num,
       replacement = "2")

# Eyelid myoclonia = 3
dfHGMD_final$Reported.phenotype.num <-
  gsub("^Eyelid.*",
       dfHGMD_final$Reported.phenotype.num,
       replacement = "3")

# Intellectual disability = 4
dfHGMD_final$Reported.phenotype.num <-
  gsub("^Intellectual.*|.*intellectual disability.*",
       dfHGMD_final$Reported.phenotype.num,
       replacement = "4")
```

For Variant Class the data will change to:
DM = 1
DM? = 2
```{r Variantclass.num}
# Make Variantclass numeric
# Assign number to Variantclass.num
dfHGMD_final$Variantclass.num<-dfHGMD_final$Variantclass
dfHGMD_final$Variantclass.num<-ifelse(dfHGMD_final$Variantclass.num=="DM", 1, 2)
```

For Overall.consequence the data will change to:
LoF = 1
nLoF = 2
```{r Overall.consequence.num}
# Make Overall.consequence numeric
# Assign number to Overall.consequence.num
dfHGMD_final$Overall.consequence.num <-
dfHGMD_final$Overall.consequence
dfHGMD_final$Overall.consequence.num <-
ifelse(dfHGMD_final$Overall.consequence.num == "LoF", 1, 2)
```

For Mutation.type the data will change to:
frameshift = 1
Missense = 2
Nonsense = 3
splice = 4
```{r Mutation.type.num}
# Make Reported.phenotype numeric
# Assign number to Reported.phenotype
# Create column
dfHGMD_final$Mutation.type.num <-
  dfHGMD_final$Mutation.type

# Use regex to make data numeric
# frameshift = 1
dfHGMD_final$Mutation.type.num <-
  gsub("^frameshift",
       dfHGMD_final$Mutation.type.num,
       replacement = "1")

# Missense = 2
dfHGMD_final$Mutation.type.num <-
  gsub("^Missense",
       dfHGMD_final$Mutation.type.num,
       replacement = "2")

# Nonsense = 3
dfHGMD_final$Mutation.type.num <-
  gsub("^Nonsense",
       dfHGMD_final$Mutation.type.num,
       replacement = "3")

# Splice = 4
dfHGMD_final$Mutation.type.num <-
  gsub("^splice",
       dfHGMD_final$Mutation.type.num,
       replacement = "4")
```

Change data to numeric and make new dataframe:
```{r shape data}
# Convert to numeric
dfHGMD_final[8] <- sapply(dfHGMD_final[8], as.numeric)
dfHGMD_final[14:18] <- sapply(dfHGMD_final[14:18], as.numeric)


# Columns to keep
keep <-
  c(
    "Consequence.num",
    "Reported.phenotype.num",
    "Variantclass.num",
    "Overall.consequence.num",
    "Mutation.type.num",
    "Year")

# Make new dataframe with keep data
data <- dfHGMD_final[keep]

# Check dataframe
str(data)
summary(data)
```
## Time series analysis in R

The ts() function will convert a numeric vector into an R time series object. The format is ts(vector, start=, end=, frequency=) where start and end are the times of the first and last observation and frequency is the number of observations per unit time (1=annual, 4=quartly, 12=monthly, etc.).   
```{r}
HGMD <- count(data, c("Year"))
HGMD

# set the freq parameter to 1 to indicate annual readings
# ts() function to create a new time series
HGMD_timeseries <- ts(HGMD$freq, start = c(min(data$Year), 1), end = c(max(data$Year), 1), frequency = 1)   
plot(HGMD_timeseries, xlab='Number of Publications for CHD2' ,ylab='Years', main='All Publications for CHD2')
## -------------   USE ARIMA MODEL  ---------------------
#creating ranges of possible values for the order parameters p, d, and q.
d <- 0 : 1
p <- 0 : 1
q <- 0 : 1
HGMD_models <- expand.grid(d = d, p = p, q = q)
head(HGMD_models, n = 4)
getTSModelAIC <- function(ts_data, p, d, q) {
                                            ts_model <- arima(ts_data, order = c(p, d, q))
                                              return(ts_model$aic)
                                              }


getTSModelAICSafe <- function(ts_data, p, d, q) {  
                                                 result = tryCatch({
                                                 getTSModelAIC(ts_data, p, d, q)
                                                  }, error = function(e) {
                                                 Inf
                                                 })
                                                 }
 
 # PICK THE BEST MODEL THAT HAS THE SMALLEST AIC 
HGMD_models$aic <- mapply(function(x, y, z) 
                              getTSModelAICSafe(HGMD_timeseries, x, y, z), HGMD_models$p, 
                              HGMD_models$d, HGMD_models$q)

subset(HGMD_models,aic == min(aic))
# ARIMA model for best p,d,q order model 
HGMD_model <- arima(HGMD_timeseries, order = c(1, 1, 1))
summary(HGMD_model)
#-------------------       Prediction    ---------------------------------
plot(forecast(HGMD_model, 1))
```

I used a frequency of 1, to look at the data annually. I did this because I only have the publication year not the month and year. The data looks good, I played the the parameters so it would not produce NANs and produce a low AIC. This model predicts that there will be another 5 publications for CHD2 next year, based off of past performance.
