---
title: "M1L2 Homework Assignment"
author: "Joshua Conte"
date: "September 17, 2017"
output: 
  pdf_document:
    fig_caption: true
header-includes:
    - \usepackage{fancyhdr}
    - \pagestyle{fancy}
    - \fancyhead[CO,CE]{DA5030}
    - \fancyfoot[CO,CE]{Assignment M1L2 by Joshua Conte}
    - \fancyfoot[LE,RO]{\thepage}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
&nbsp;


\clearpage
\setcounter{secnumdepth}{5}
\setcounter{tocdepth}{5}
\pagebreak

# M1L2 Homework Assignment
R studio was configured with the following parameters before beginning the project:
```{r Configure RStudio, warning = FALSE, message = FALSE}
# clears the console in RStudio
cat("\014") 

# clears environment
rm(list = ls())  

# Set working directory
setwd("C:/R/DA5030/module_01")

# Load required packages
require(ggplot2)
require(reshape2)
library(knitr) # Used for making tables in this report
```
## Part 1: Cauchy distributions
From Wikipedia, the Cauchy distribution, named after Augustin Cauchy, is a continuous probability distribution. It is also known, especially among physicists, as the Lorentz distribution (after Hendrik Lorentz), Cauchy-Lorentz distribution, Lorentz(ian) function, or Breit-Wigner distribution. The Cauchy distribution is the distribution of the x-intercept of a ray issuing from with a uniformly distributed angle. It is also the distribution of the ratio of two independent normally distributed random variables if the denominator distribution has mean zero.

### Cauchy Probability Density Function
I used the Cauchy distribution feature in R.  dcauchy is used for density location is x0 and scale is gamma. I used the same location and scale parameters in the Wikipedia example.

For the ggplot info, I used lower limit of -5 and upper of 5. I used fun and args for the function and aes for aesthetic mappings. I added theme to make the plot more aesthetic too.
```{r Cauchy Probability Density Function}
# Cauchy Probability Density Function:

x_lower <- -5
x_upper <- 5

ggplot(data.frame(x = c(x_lower , x_upper)), aes(x = x)) +
  xlim(c(x_lower , x_upper)) +
  stat_function(fun = dcauchy,
                args = list(location = 0, scale = 0.5),
                aes(colour = "0 & 0.5")) +
  stat_function(fun = dcauchy,
                args = list(location = 0, scale = 1),
                aes(colour = "0 & 1")) +
  stat_function(fun = dcauchy,
                args = list(location = 0, scale = 2),
                aes(colour = "0 & 2")) +
  stat_function(fun = dcauchy,
                args = list(location = -2, scale = 1),
                aes(colour = "-2 & 1")) +
  scale_color_manual("Location & Scale \n Parameters",
                     values = c("green", "orange", "purple", "light blue")) +
  labs(x = "\n x", y = "P(x) \n",
       title = "Cauchy Probability Density Function") +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.title.x = element_text(
      face = "bold",
      colour = "blue",
      size = 12
    ),
    axis.title.y = element_text(
      face = "bold",
      colour = "blue",
      size = 12
    ),
    legend.title = element_text(face = "bold", size = 8),
    legend.position = "right"
  )
```

### Cauchy Cumulative Distribution Function
This is the same as above but pcauchy is used for distribution function. 
```{r Cauchy Cumulative Distribution Function}
# Cauchy Cumulative Distribution Function:

x_lower <- -5
x_upper <- 5

ggplot(data.frame(x = c(x_lower , x_upper)), aes(x = x)) +
  xlim(c(x_lower , x_upper)) +
  stat_function(fun = pcauchy,
                args = list(location = 0, scale = 0.5),
                aes(colour = "0 & 0.5")) +
  stat_function(fun = pcauchy,
                args = list(location = 0, scale = 1),
                aes(colour = "0 & 1")) +
  stat_function(fun = pcauchy,
                args = list(location = 0, scale = 2),
                aes(colour = "0 & 2")) +
  stat_function(fun = pcauchy,
                args = list(location = -2, scale = 1),
                aes(colour = "-2 & 1")) +
  scale_color_manual("Location & Scale \n Parameters",
                     values = c("green", "orange", "purple", "light blue")) +
  labs(x = "\n x", y = "P(x) \n",
       title = "Cauchy Cumulative Distribution Function") +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.title.x = element_text(
      face = "bold",
      colour = "blue",
      size = 12
    ),
    axis.title.y = element_text(
      face = "bold",
      colour = "blue",
      size = 12
    ),
    legend.title = element_text(face = "bold", size = 8),
    legend.position = "right"
  )
```

## Part 2: M01_Lesson_02_Q1
### load the file M01_Lesson_02_Q1.csv.   
I opened the data using read.csv2 and removed the first column since it appears to be row numbers.  I also changed the data to numeric so I could analyze it with R.

Below is my R code:
```{r open csv file}
# Some VCF files are really big and take a while to open.  This command checks to
# see if it is already opened, if it is, it does not open it again.
# I also omitted the first column
if (!exists("csv.df")) {
  csv.df <-
    read.csv2(
      'M01_Lesson_02_Q1.csv',
      sep = ",",
      stringsAsFactors = FALSE,
      row.names = NULL,
      header = TRUE,
      colClasses = c("NULL", NA, NA, NA, NA, NA)
    )
}

# Check to make sure the data is in numeric form for analysis:
sapply(csv.df, class)

# It looks like the data is in character and integer form and I will change it all
# to numeric for easy analysis:
df1 <-
  data.frame(sapply(csv.df, function(x)
    as.numeric(as.character(x))))

# Check to make sure it is in numeric form:
sapply(df1, class)
```

### How is the data distributed?
I concluded that:

* column A is Normal
* column B is Normal
* column C is Poisson
* column D is Poisson
* column E is Uniform

My analysis is listed below:

To understand how the data is distributed, I needed to review the data and plot it too. I began with a Shapiro test to see if the P value was greater than 0.05 to confirm that we cannot reject the hypothesis that the sample comes from a population which has a normal distribution. Then use QQ plots to verify the results:
```{r data distributed analysis 1}

# I begain by running a Shapiro-Wilk test:
shapiro.test(df1$A)
shapiro.test(df1$B)
shapiro.test(df1$C)
shapiro.test(df1$D)
shapiro.test(df1$E)

# Then I plotted the data:
qqnorm(df1$A)
qqline (df1$A, col=2)

qqnorm(df1$B)
qqline (df1$B, col=2)

qqnorm(df1$C)
qqline (df1$C, col=2)

qqnorm(df1$D)
qqline (df1$D, col=2)

qqnorm(df1$E)
qqline (df1$E, col=2)
```

Looking at the results of the Shapiro-Wilk test, regarding A and B, we cannot reject the hypothesis that the sample comes from a population which has a normal distribution.  For C, D, and E we can reject the hypothesis that the sample comes from a population which has a normal distribution.  Verifying the plots, A and B are normal and the others are not.

Regarding C, D , and E, I needed to run some more tests using fitdistrplus:
```{r data distributed analysis 2}
library(fitdistrplus)

# This plots the data
descdist(df1$C, discrete = FALSE)
descdist(df1$D, discrete = FALSE)
descdist(df1$E, discrete = FALSE)
```
E is uniform (see below too), but C and D require additional tests but will be binomial or Poisson.

Based off of the fitdist plots, C is Poisson and D is binomial:
```{r data distributed analysis 3}
# For C
fit.pois2<-fitdist(df1$C, "pois")
plot(fit.pois2)

# For D
fit.pois<-fitdist(df1$D, "pois")
plot(fit.pois)

# For E
fit.unif<-fitdist(df1$E, "unif")
plot(fit.unif)
```


### What are the summary statistics?
Below are the summary statistics with the graphs:
```{r data distributed}
# internal structure 
str(df1)

# This is a summary of the data:
summary(df1)

# Melt the data requires reshape2
rnd <- melt(data = df1)

# Review the data to make sure it looks ok
summary(rnd)

# Plot the data:
ggplot(rnd, aes(x = value)) + geom_density(aes(
  group = variable,
  color = variable,
  fill = variable
)) + labs(title = "M01_Lesson_02_Q1.csv Data", y = "??,sigma^2(X)", x =
            " successes")

# The data looks hard to read, so I am going to separate A-D and E.
# This is A-D:
df2 <- data.frame(
  A = df1$A,
  B = df1$B,
  C = df1$C,
  D = df1$D
)
rnd2 <- melt(data = df2)
summary(rnd2)

ggplot(rnd2, aes(x = value)) + geom_density(aes(
  group = variable,
  color = variable,
  fill = variable
)) + labs(title = "M01_Lesson_02_Q1.csv Data for A - D", y = "??,sigma^2(X)", x =
            " successes")

# This is E:
df3 <- data.frame(E = df1$E)
rnd3 <- melt(data = df3)
summary(rnd3)

ggplot(rnd3, aes(x = value)) + geom_density(aes(
  group = variable,
  color = variable,
  fill = variable
)) + labs(title = "M01_Lesson_02_Q1.csv Data for E", y = "??,sigma^2(X)", x =
            " successes")

```

### Are there anomalies/outliers?
I had a hard time trying to understand if C and D were possion or binomial distributed, other than that I do not think there are any anomalies and the data looks good.

### Try to regenerate the data in each column and plot your regenerated data versus the original data using a faceted graph. How does it compare?
Below is regeneration of the data using R:
```{r data distributed regenerate}
n = 333
norm_dist <- data.frame(
  regenerateA = rnorm(n = n, mean = mean(df1$A), sd = sd(df1$A)),
  regenerateB = rnorm(n = n, mean = mean(df1$B), sd = sd(df1$B)),
  regenerateC = rpois(n = n, lambda = mean(df1$C)), 
  regenerateD = rpois(n = n, lambda = mean(df1$D)),
  regenerateE = runif(n, min(df1$E), max(df1$E)+1)
)

summary(norm_dist)
head(norm_dist)

# Melt the data
rndReg <- melt(data = norm_dist)

# Review the data to make sure it looks ok
summary(rndReg)

# Plot the data:
ggplot(rndReg, aes(x = value)) + geom_density(aes(
  group = variable,
  color = variable,
  fill = variable
)) + labs(title = "Regenerate Data", y = "??,sigma^2(X)", x =
            " successes")
```

Comparing the data one to one, it looks like it matches well.  However, I had a hard time getting C to match up, it looks like it could be in the ballpark but it doesn't fit as well as the others:
```{r data distributed analysis A}
distA <- data.frame(A = df1$A, regenerateA = norm_dist$regenerateA)
distA <- melt(data = distA)
ggplot(distA, aes(x = value)) + geom_density(aes(
  group = variable,
  color = variable
)) + labs(title = "A vs regenerateA", y = "??,sigma^2(X)", x =
            " successes")
```

```{r data distributed analysis B}
distB <- data.frame(B = df1$B, regenerateB = norm_dist$regenerateB)
distB <- melt(data = distB)
ggplot(distB, aes(x = value)) + geom_density(aes(
  group = variable,
  color = variable
)) + labs(title = "B vs regenerateB", y = "??,sigma^2(X)", x =
            " successes")
```

```{r data distributed analysis C}
distC <- data.frame(C = df1$C, regenerateC = norm_dist$regenerateC)
distC <- melt(data = distC)
ggplot(distC, aes(x = value)) + geom_density(aes(
  group = variable,
  color = variable
)) + labs(title = "C vs regenerateC", y = "??,sigma^2(X)", x =
            " successes")
```

```{r data distributed analysis D}
distD <- data.frame(D = df1$D, regenerateD = norm_dist$regenerateD)
distD <- melt(data = distD)
ggplot(distD, aes(x = value)) + geom_density(aes(
  group = variable,
  color = variable
)) + labs(title = "D vs regenerateD", y = "??,sigma^2(X)", x =
            " successes")
```

```{r data distributed analysis E}
distE <- data.frame(E = df1$E, regenerateE = norm_dist$regenerateE)
distE <- melt(data = distE)
ggplot(distE, aes(x = value)) + geom_density(aes(
  group = variable,
  color = variable
)) + labs(title = "M01_Lesson_02_Q1.csv Data for E", y = "??,sigma^2(X)", x =
            " successes")
```
